Assessment of forestry road network projection methods using the “roads” package in R

### 2022-10-06 analysis/paper/paper.Rmd
re-run projections and updates to the paper. WIP waiting on Hardy results at 200m

@@

To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster. We used several spatial raster layers to construct our cost surface. For all cells the base cost followed the Interior Appraisal Manual of BC where the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a]. Separate costs for river and pipeline crossings were then added to the cost. Major water bodies were considered barriers and assigned a very high cost. Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned a cost of 0 (Table 1). The cost surface for the Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, and minimum spanning trees. Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two. The path to each cutblock is determined independently of the other cutblocks. The cost is represented by a cost raster described below. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the LCP and MST approaches given their assumptions are better aligned with how road networks are designed.

In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>). This plugin uses a similar least cost path algorithm to the `roads` package but instead of building roads along the least cost path to each cutblock independently the Forest Road Network Plugin uses a heuristic to determine which cutblocks to connect first and then builds the roads sequentially. The default heuristic is to connect the closest cutblocks first. We supplied our cost surface, cutblocks and road network data and kept all the default settings within the QGIS plugin.

@@

We projected road development in the Revelstoke TSA from 1990 to the present using several different inputs and compared the projections to the observed present day road network in order to evaluate the accuracy of the projections and their sensitivity to different inputs. We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)). Note that the default skidding distance used by the Hardy QGIS plugin is 100 m which corresponds to 25 pts/km^2^ .

@@

library(raster)
library(dplyr)
library(sf)
library(roads)
library(terra)
library(tmap)
library(purrr)
library(caribouMetrics)
library(pfocal)

@@

# read in example area data
exRoads <- read_sf(here(data_path_drvd, "..", "testing_ex_roads.gpkg"))
tsb <- read_sf(here(data_path_drvd, "..", "testing_tsb.gpkg"))
cutblocks <- read_sf(here(data_path_drvd, "..", "testing_cutblocks.gpkg"))
tsaCost <- raster(here(data_path_drvd, "..", "testing_cost.tif"))
roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))

@@

```{r fig3, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

>>>>

devtools::load_all(here())

@@

To spatially define the cost of constructing and traveling on a forestry road we developed a cost raster. We used several spatial raster layers to construct our cost surface. Following the Interior Appraisal Manual of BC, the base the cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a]. Separate costs for river and pipeline crossings were then added and major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0 (Table 1). The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

@@

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke TSA [@teucher2021].

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees. Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two. The path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is detemined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed.

In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>). This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks.

@@

We projected road development in the Revelstoke TSA from 1990 to the present using several different inputs and compared the projections to the observed present day road network in order to evaluate the accuracy of the projections and their sensitivity to different inputs. We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by comparing using the polygon centroid and regularly spaced and random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)). For the Hardy QGIS plugind supplied our cost surface, cutblocks, road network data and kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^ .

@@

library(dplyr)
library(sf)
library(roads)
library(terra)
library(tmap)
library(purrr)
library(caribouMetrics)
library(pfocal)

@@

# read in example area data
exRoads <- read_sf(here(data_path_drvd, "..", "testing_ex_roads.gpkg"))
tsb <- read_sf(here(data_path_drvd, "..", "testing_tsb.gpkg"))
cutblocks <- read_sf(here(data_path_drvd, "..", "testing_cutblocks.gpkg"))
tsaCost <- raster::raster(here(data_path_drvd, "..", "testing_cost.tif"))
roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))

# densities
high <- 0.00001
low <-  0.000001

@@

```{r fig3, fig.height=4.5, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)
